An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-11-12 paper/main.tex
weiter

Researchers have used a range of techniques, like gradient descent, heuristics,
and even evolutionary algorithms, to find the best bitwidth settings for each
layer. \\\\ Another method that helps reduce resource usage is network pruning.
This means cutting out parts of the network that do not contribute much to its
performance. A way to do this is simulated annealing, a search algorithm that
mimics the process of cooling metal to find a stable structure. It does not
need gradient information and can still find good network shapes by exploring
many possibilities. This makes it useful when you want to shrink a network
without going through heavy retraining. \\\\ Previous studies have shown that
reducing the precision of neural network parameters can lead to significant
improvements in memory usage and processing speed. Even when precision is
reduced to very low levels, models can still perform competitively on complex
tasks. However, while these early results are promising, the overall effects of
lowering precision, especially how it influences model accuracy, training
behavior, and storage efficiency, are still not fully understood and deserve
more detailed investigation. \\\\ This is where our experimental work begins.
In this paper, we take a closer look at how different levels of parameter
precision affect the behavior and performance of neural networks. Our goal is
to understand what happens when we reduce precision in various ways, and how
much we can lower it before performance starts to drop. \\\\ We start by
testing how reducing precision during both training and inference affects model
accuracy and runtime. Next, we look at what happens when we train the network
using full precision, but lower the precision only during inference. This setup
is common in real-world applications, where a powerful machine handles
training, but the final model is deployed on a lightweight device. Then, we run
a series of tests where we round the trained weights to a limited number of
decimal places. This helps to see how much we can simplify the stored weight
values without hurting accuracy too much. This is especially useful for devices
with very limited storage. We also try out mixed-precision training, using a
mix of Half-Precision Floating Point (FP16) and Single-Precision Floating Point
(FP32) formats. On top of that, we experiment with different types of gradient
clipping, like global norm clipping and value clipping, to see how they affect
training stability and final accuracy. \\\\ Even though we use relatively
simple network architectures and datasets, our results show that neural
networks are quite robust to low-precision settings. In some cases, we were
able to round weights to just three decimal digits and still get nearly the
same accuracy as the full-precision version. These findings show that it is
possible to significantly reduce the memory and compute needs of neural
networks without losing much performance. This makes them more practical for
use in low-resource settings. It also opens the door for more advanced
techniques that combine quantization, pruning, and precision-aware training to
build highly efficient models for real-world applications.

@@

We trained a simple fully connected neural network on the MNIST dataset,
consisting of 60,000 training and 10,000 test grayscale images of handwritten
digits (28×28 pixels, see figure~\ref{fig:mnistSample}). Model training was
performed under varying parameter bitwidths, and 5-fold cross-validation to
evaluate the model performance. The structure of the MLP is depicted in
figure~\ref{fig:mlpStructure}.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/mlpStruc.png}
	\caption{MLP structure used in the experiments.}\label{fig:mlpStructure}
\end{figure}

>>>>

Researchers have used a range of techniques, like stochastic gradient descent
(SGD), heuristics, and even evolutionary algorithms, to find the best bitwidth
settings for each layer. \\\\ Another method that helps reduce resource usage
is network pruning. This means cutting out parts of the network that do not
contribute much to its performance. A way to do this is simulated annealing, a
search algorithm that mimics the process of cooling metal to find a stable
structure. It does not need gradient information and can still find good
network shapes by exploring many possibilities. This makes it useful when you
want to shrink a network without going through heavy retraining. \\\\ Previous
studies have shown that reducing the precision of neural network parameters can
lead to significant improvements in memory usage and processing speed. Even
when precision is reduced to very low levels, models can still perform
competitively on complex tasks. However, while these early results are
promising, the overall effects of lowering precision, especially how it
influences model accuracy, training behavior, and storage efficiency, are still
not fully understood and deserve more detailed investigation. \\\\ This is
where our experimental work begins. In this paper, we take a closer look at how
different levels of parameter precision affect the behavior and performance of
neural networks. Our goal is to understand what happens when we reduce
precision in various ways, and how much we can lower it before performance
starts to drop. \\\\ We start by testing how reducing precision during both
training and inference affects model accuracy and runtime. Next, we look at
what happens when we train the network using full precision, but lower the
precision only during inference. This setup is common in real-world
applications, where a powerful machine handles training, but the final model is
deployed on a lightweight device. Then, we run a series of tests where we round
the trained weights to a limited number of decimal places. This helps to see
how much we can simplify the stored weight values without hurting accuracy too
much. This is especially useful for devices with very limited storage. We also
try out mixed-precision training, using a mix of Half-Precision Floating Point
(FP16) and Single-Precision Floating Point (FP32) formats. On top of that, we
experiment with different types of gradient clipping, like global norm clipping
and value clipping, to see how they affect training stability and final
accuracy. \\\\ Even though we use relatively simple network architectures and
datasets, our results show that neural networks are quite robust to
low-precision settings. In some cases, we were able to round weights to just
three decimal digits and still get nearly the same accuracy as the
full-precision version. These findings show that it is possible to
significantly reduce the memory and compute needs of neural networks without
losing much performance. This makes them more practical for use in low-resource
settings. It also opens the door for more advanced techniques that combine
quantization, pruning, and precision-aware training to build highly efficient
models for real-world applications.

@@

The effect of numerical precision on training and inference is evaluated
through five experiment groups run on standard image classification tasks. The
main dataset MNIST\footnote{LeCun, Cortes, and Burges:
	\url{http://yann.lecun.com/exdb/mnist/}.} ($28\times 28$ grayscale images of
handwritten digits, train/test split $60.000$/$10.000$, 10 classes, examples
shown in Figure~\ref{fig:mnistSample}) is used to determine precision effects
on accuracy and runtime under a simple workload, while an after-training
rounding experiment is performed on FashionMNIST\footnote{Xiao, Rasul, and
	Vollgraf: \url{https://github.com/zalandoresearch/fashion-mnist}.} ($28\times
28$ grayscale images of clothing items, train/test split $60.000$/$10.000$, 10
classes) so the impact of using coarse value sets is better observable.\\

@@

The classifier used is a fully connected network
$784\rightarrow128\rightarrow64\rightarrow10$ with ReLU between linear layers
(Figure~\ref{fig:mlpStructure}). Inputs are normalized with mean $0.1307$ and
std $0.3081$. Unless stated otherwise, we train with cross-entropy loss and SGD
(learning rate $0.01$, momentum $0.9$), batch size $64$, and on CUDA.\\

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/mlpStruc.png}
	\caption{Network structure used in the experiments.}\label{fig:mlpStructure}
\end{figure}
